---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# crone

<!-- badges: start -->
<!-- badges: end -->

The goal of *crone* is to carry out most of the task and calculations of structural crystallography using 1D crystal structure models. The 1D models included with the package are taken from existing linear chemical molecules. A full description of the theory behind 1D crystallography and the *crone* package can be studied in this [article](https://iopscience.iop.org/article/10.1088/1361-6404/aa8188). The package is aimed at undergraduate and graduate students that need to be trained in structural crystallography or in the development of methods for structural crystallography.

## Installation

You can install the released version of crone from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("crone")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jfoadi/crone")
```
## Example

Structures are stored in named lists of type `sdata`:

```{r example1}
library(crone)

# Load carbon dioxide
sdata <- load_structure("carbon_dioxide")

# sdata includes: 
# cell length(a), space group (SG),atoms position (x0), 
# atomic numbers (Z),B-factors (B), occupancies (occ)
sdata
```

Once a structure is loaded (or created from scratch), the ideal electron density can be calculated and plotted:

```{r example2}
rtmp <- structure_gauss(sdata=sdata,N=1000)
plot(rtmp$x,rtmp$rr,type="l",
     xlab=expression(x),ylab=expression(rho))
```

The repeated motive in the unit cell can be plotted for several unit cells:

```{r example3}
x <- seq(-sdata$a*10,sdata$a*10,length=20000)
rtmp <- structure_gauss(sdata=sdata,x=x)
plot(rtmp$x,rtmp$rr,type="l",
     xlab=expression(x),ylab=expression(rho))
```

A diffraction pattern due to just one unit cell or to multiple unit cells is created with function `diffraction`:

```{r example4}
# One unit cell
dtmp <- diffraction(sdata=sdata,D=1,Ncell=1)
plot(dtmp$xstar,dtmp$Imod,type="l",xlab=expression(s),
     ylab=expression(I))

# 10 unit cells
dtmp <- diffraction(sdata=sdata,D=1,Ncell=10)
plot(dtmp$xstar,dtmp$Imod,type="l",xlab=expression(s),
     ylab=expression(I))
```

Many more examples can be studied in the five tutorials included as package vignettes. The first three tutorials are set at an introductory/intermediate level; the last two tutorials are more difficult to digest.

## Reference

Details of the theory and algorithms behind ```crone``` can be found in: E. Smith, G. Evans and J. Foadi (2017), "An effective introduction to structural crystallography using 1D Gaussian atoms", Eur. J. Phys. 38, [link](https://doi.org/10.1088/1361-6404/aa8188)
