---
title: "Read and analyse an MTZ file"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Read and analyse an MTZ file}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<!-- Block with global markdown options
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
-->

## Introduction
Main aim of this tutorial is to load the full content of an MTZ file in the workspace, to analyse and modify data, and to write the modified content into a new MTZ file.

## Sample MTZ file
Some sample files are stored as external data in this package. These are the MTZ files available with the current release. To access the file, first load the `cry` package.

```{r, ech=TRUE}
library(cry)
```

Next, have a look at what is included in the external-data directory of `cry`.

```{r echo=TRUE}
datadir <- system.file("extdata",package="cry")
all_files <- list.files(datadir)
print(all_files)
```

We are interested in the MTZ file "1dei_phases.mtz". This file can be loaded using function `readMTZ`.

## The structure of an MTZ file in R
Let's load "1dei_phases.mtz" in memory. The created object is a named list.

```{r echo=TRUE}
filename <- file.path(datadir,"1dei_phases.mtz")
objMTZ <- readMTZ(filename)
class(objMTZ)
names(objMTZ)
```
Each component of \code{objMTZ} is either a named list, or a data frame. More specifically:

1. `reflections`


   It's a data frame. It contains the actual x-ray diffraction 
   data from a given crystal.
   
2. `header`


   It's a named list. It contains information on the crystal 
   used to obtain the x-ray diffraction data and on the 
   diffraction experiment, in general.
   
3. `batch_header`
   It's a named list. It gets filled only when the MTZ file       
   includes so-called "unmerged" data. 
   
   
For the merged MTZ file we are exploring in this tutorial, the `batch_header` is `NULL`.

## The MTZ `header`
This is a named list. The names hint at the content of the specific components.

```{r echo=TRUE}
hdr <- objMTZ$header
class(hdr)
names(hdr)
```

For example, `NCOL` is the number of columns of the `reflections` data frame, `CELL` contains the crystal unit cell parameters and `SYMM` includes information on the crystal symmetry.

A merged MTZ file stores reflections x-ray data in a hierarchical way. The main purpose of collecting x-ray data is to determine the 3D structure of the molecule arranged in an ordered crystallographic lattice. Reflections related to a same molecule can come from different crystals. Different datasets, corresponding to different rotation sweeps from the same crystal, can be related to a crystal. Each MTZ file can thus include data from several crystals, each crystal giving origin to several datasets. Normally, a single MTZ file will be part of a same project, but this does not necessarily have to be the case. In summary, the hierarchy defining each dataset is:
<center>
 *Project* $\quad\Rightarrow\quad$ *Crystal* $\quad\Rightarrow\quad$ *Dataset*
</center>
For the file loaded we find:

```{r echo=TRUE}
print(objMTZ$header$PROJECT)
print(objMTZ$header$CRYSTAL)
print(objMTZ$header$DATASET)
```

The only dataset in this MTZ file is called *data_1*. It was obtained from x-ray diffraction of a crystal called *cryst_1*. This crystal is one of the samples belonging to a project called *sf_convert*. All MTZ files contain, by default, a dataset *HKL_base* which comes from the crystal *HKL_base*, which is part of the project *HKL_base*; this peculiar project is always present to make sure that data observations made only of the Miller indices, are always included. More information on the header's content can be read in the documentation for `readMTZ` or `readMTZHeader`.

## The MTZ `reflections`
Data are actually contained in this component. It is a data frame whose columns have the names found in the header component `COLUMN[,1]`:

```{r echo=TRUE}
objMTZ$header$COLUMN[,1]
objMTZ$reflections[1:5,]
```

A data frame is the appropriate class for observations like the crystallographic x-ray data. One reason for this is, for instance, that grouping or selection according to a specific criterion can be carried out very easily with a data frame. Let us, as an example, select part of the data using a condition on the Miller indices. We can be interested to select reflections for which $h$ has a specific value; this is shown in the following chunk of code.

```{r echo=TRUE}
# List the different values of H
unique(objMTZ$reflections$H)

# Select all reflections with H=1
idx <- which(objMTZ$reflections$H == 1)
length(idx)  # 373 reflections have H=0

# Save these reflections in a different object
refs <- objMTZ$reflections[idx,]

# Show the first 10 selected reflections
refs[1:10,]

# Find out the range of FP/sigFP for the selected reflections
range(refs$FP/refs$SIGFP)
```

Clearly, a lot of different operations and investigations can be carried out on the selected data. A second example involves data for which the signal-to-noise ratio (FP/sigFP) is greater than 1, as shown in the following snippet.

```{r echo=TRUE,fig.height=3,fig.width=5}
idx <- which(objMTZ$reflections$FP/objMTZ$reflections$SIGFP >= 1)
length(idx) # 8377 reflections have FP/sigFP >= 1

# The reflections can be extracted
refs <- objMTZ$reflections[idx,]

# Histogram of I/sigI
hist(refs$FP/refs$SIGFP,breaks=30,
     main="FP/sigFP",xlab="FP/sigFP")
```

A third example is related to the reflections' resolution. Once it has been calculated (cell parameters are needed to do that), data can be selected within a given resolution range (a so-called *resolution shell*). There is a function in `cry`, called `hkl_to_reso`, which calculates the resolution in angstroms for a reflection of Miller indices $h, k, l$.

```{r echo=TRUE}
# Extract cell parameters from header
cpars <- objMTZ$header$CELL
print(cpars)
a <- cpars[1]; b <- cpars[2]; c <- cpars[3]
aa <- cpars[4]; bb <- cpars[5]; cc <- cpars[6]

# Resolution of reflection (1,0,0) (0,1,0) (0,0,1)
hkl_to_reso(1,0,0,a,b,c,aa,bb,cc)
hkl_to_reso(0,1,0,a,b,c,aa,bb,cc)
hkl_to_reso(0,0,1,a,b,c,aa,bb,cc)

# Reflections with higher Miller indices have higher resolutions
hkl_to_reso(10,0,0,a,b,c,aa,bb,cc)
hkl_to_reso(10,0,-20,a,b,c,aa,bb,cc)
```

Once resolutions are calculated (and the calculation is done on data having the same order as the original reflections), data within a given resolution shell can be selected.

```{r echo=TRUE}
resos <- hkl_to_reso(objMTZ$reflections$H,
                     objMTZ$reflections$K,
                     objMTZ$reflections$L,
                     a,b,c,aa,bb,cc)

# Resolution range for all data
range(resos)

# Select data with resolution between 5 and 9 angstroms
idx <- which(resos >= 5 & resos <= 9)
length(idx)  # Only 314 reflections
```

## Output to a new MTZ file
In order to investigate specific ideas it might be worth to modify the observations in a specific way and to write them out to a new MTZ file to be later handled by the tools proper of the [CCP4](http://www.ccp4.ac.uk) family of programs for crystallography. The `cry` function to write reflections content to an MTZ file is called `writeMTZ`. In order to make this a flexible function, it has been deemed appropriate to take as input the three named lists returned by `readMTZ`. These will have been changed by the user, prior to use with `writeMTZ`. By default, the MTZ file title is left unchanged and the `batch_header` list is set to `NULL`; thus by default the MTZ file is assumed to be a merged-reflections file. Special attention will have to be devoted to parts of the `header` list, which store information on the observations. These are, for instance, `NCOL`, `SORT`, `RESO`, `COLUMN`, etc. An example can help to clarify this concept.

A typical modifications of the observed data is when high-resolution reflections are eliminated. This means that the number of reflections and other quantities in the `header` will have to be changed.

```{r echo=TRUE}
# Copy original data to 2 separate lists
refs <- objMTZ$reflections
hdr <- objMTZ$header
length(refs[,1])  # Number of reflections before cut
print(hdr$NCOL[2])

# Cut data to 5 angstroms resolution
# (see previous chunk of code for resos)
idx <- which(resos >= 5)
refs <- refs[idx,]
length(refs[,1])  # Now there are only 333 observations

# Modify specific parts of header
hdr$NCOL[2] <- length(refs[,1])            # Number of reflections
hdr$RESO <- sort(1/range(resos[idx])^2)    # Resolution range
hdr$DATASET[2,2] <- "Data cut to 5A"  # Dataset
obsmin <- apply(refs,2,min,na.rm=TRUE)
obsmax <- apply(refs,2,max,na.rm=TRUE)
hdr$COLUMN[,3] <- obsmin                   # COLUMN
hdr$COLUMN[,4] <- obsmax
```

The data frame `COLSRC` contains the date and time at which the specific data columns have been generated. This can be changed
with the `cry` function `change_COLSRC`; the new date and time is
the current one.

```{r echo=TRUE}
# Original COLSRC
print(hdr$COLSRC)

# Change date and time
hdr <- change_COLSRC(hdr)

# New COLSRC
print(hdr$COLSRC)
```

Once all changes have been done, the list components are ready to
be saved out to a new MTZ file called, in this specific instance,
"new.mtz".

```{r echo=TRUE}
# Temporary directory for output
tdir <- tempdir()
fname <- file.path(tdir,"new.mtz")

# Write changed data to the new MTZ file
writeMTZ(refs,hdr,fname,title="New truncated dataset")
```

Data from the new MTZ file can either be explored using CCP4
programs, or read back into R using again `readMTZ`.

```{r echo=TRUE}
# Read data from "new.mtz"
newMTZ <- readMTZ(fname,TRUE)
```
